---
author: Cathy Stanton
categories: blog-posts
date: "2024-09-09"
slug: week-1-9-09
title: Blog Post\: Week of 9/09/2024
---

#' @title GOV 1347: Introductory Blog Post/Laboratory Session
#' @author Matthew E. Dardet
#' @date September 4, 2024

####----------------------------------------------------------#
#### Preamble
####----------------------------------------------------------#

# Load libraries.
## install via `install.packages("name")`
```{r}
library(ggplot2)
library(maps)
library(tidyverse)
```

## set working directory here
# setwd("~")

####----------------------------------------------------------#
#### Read and clean presidential popular vote.
####----------------------------------------------------------#

```{r}
# Read presidential popular vote. 
d_popvote <- read_csv("popvote_1948-2020.csv")

# Subset data to most recent past election year. 
d_popvote |> 
  filter(year == 2020) |> 
  select(party, candidate, pv2p)

# Pivot data to wide format with party names as columns and two-party vote share as values.
(d_popvote_wide <- d_popvote |>
    select(year, party, pv2p) |>
    pivot_wider(names_from = party, values_from = pv2p))

# Modify winner column to show "D" if Democrats win and "R" if Republicans win. 
(d_popvote_wide <- d_popvote_wide |> 
    mutate(winner = case_when(democrat > republican ~ "D",
                              TRUE ~ "R")))

# Summarize data with respect to winners. 
d_popvote_wide |> 
  group_by(winner) |>
  summarise(races = n())
```


####----------------------------------------------------------#
#### Visualize trends in national presidential popular vote. 
####----------------------------------------------------------#


```{r}
# Visualize the two-party presidential popular over time. 
ggplot(data = d_popvote, aes(x=year, y=pv)) +
  geom_line()
```


####----------------------------------------------------------#
#### State-by-state map of presidential popular votes.
####----------------------------------------------------------#


```{r}
# Sequester shapefile of states from `maps` library.
states_map <- map_data("state")

# Read wide version of dataset that can be used to compare candidate votes with one another. 
d_pvstate_wide <- read_csv("clean_wide_state_2pv_1948_2020.csv")

# Merge d_pvstate_wide with state_map.
d_pvstate_wide$region <- tolower(d_pvstate_wide$state)

pv_map <- d_pvstate_wide |>
  filter(year == 2020) |>
  left_join(states_map, by = "region")

# Make map grid of state winners for each election year available in the dataset. 
pv_win_map <- d_pvstate_wide %>%
  filter(year == 2020) %>%
  left_join(states_map, by="region") %>%
  mutate(winner=ifelse(R_pv > D_pv, "republican", "democrat"))

pv_win_map %>%
  ggplot(aes(long, lat, group=group)) +
  geom_polygon(aes(fill=winner), color="black") +
  scale_fill_manual(values = c("dodgerblue4", "firebrick1"))

d_pvstate_wide %>%
  filter(year >= 1980) %>%
  full_join(states_map, by="region") %>%
  mutate(winner = ifelse(R_pv > D_pv, "republican", "democrat")) %>%
  ggplot(aes(long, lat, group=group)) +
    facet_wrap(facets = year ~.) + 
    geom_polygon(aes(fill=winner), color="white") +
    scale_fill_manual(values=c("dodgerblue4", "firebrick1")) + 
    theme_void()
```


####----------------------------------------------------------#
#### Forecast: simplified electoral cycle model. 
####----------------------------------------------------------#


```{r}
# Create prediction (pv2p and margin) based on simplified electoral cycle model: 
# vote_2024 = 3/4*vote_2020 + 1/4*vote_2016 (lag1, lag2, respectively). 

d_pvstate_wide %>% colnames()

pv2p_2024_states <- d_pv_state_wide %>%
  filter(year == 2020) %>%
  group_by(state) %>%
  summarize(D_pv2p_2024 = 0.75*D_pv2p + 0.25*D_pv2p_lag1,
            R_pv2p_2024 = 0.75*R_pv2p + 0.25*R_pv2p_lag1) %>%
  mutate(pv2p_2024_margin = R_pv2p_2024 - D_pv2p_2024,
         region = lower(state))

# Plot the margin of victory in a U.S. state map.
pv2p_2024_states %>%
  left_join(states_map, by="region") %>%
  ggplot(aes(lat, long, group=group)) + 
    geom_polygon(aes(fill = pv2p_2024_margin), color="black") + 
    scale_fill_gradient2(high="firebrick1",
                         mid="white",
                         low="dodgerblue4",
                         breaks=c(-50, -25, 0, 25, 50),
                         limits=c(-50, 50)) +
    theme_void()

# Generate projected state winners and merge with electoral college votes to make 
# summary of electoral college vote distributions. 

ec <- read_csv("ec_full.csv")

pv2p_2024_states <- pv2p_2024_states %>%
  mutate(#TODO) %>%
  left_join(#TDOD)

pv2p_2024_states %>%
  group_by(winner) %>%
  summarize(#TODO)
```


## Harris: 
## Trump: 