---
title: Forecast 09/30/2024
author: Catherine Stanton
date: '2024-09-29'
slug: week-4-forecast
categories: []
tags: []
---

<h1>Week 4: September 30, 2024</h1>  
This blog is an ongoing project for Professor Ryan Enos' Election Analytics Course at Harvard College (GOV 1347, Fall 2024). It will be updated weekly with posts analyzing how different features impact the likelihood of Kamala Harris (D) or Donald Trump (R) winning the 2024 U.S. Presidential Election or winning specific states in the election. The blog will culminate in a final predictive model for the outcome of the general election.

<h2>Context & Question</h2>
In the 18 elections since the end of World War II (1948), incumbent candidates have won a third of them. But considering that the incumbent president has only run in 11 of those elections, the statistic comes out to be that the incumbent won in about **two-thirds** of elections in which they were running. So this week's blog explores whether incumbency really matters in a presidential race, and whether it's a good prediction of an election outcome. We can measure "incumbency" both on the basis of the candidate and the party. This is especially interesting in the 2024 race because one candidate (Donald Trump) is not a *direct* incumbent, meaning he did not serve in the pevious term, but he has served before. And the other candidate (Kamala Harris) is a member of the incumbent party *and* incumbent administration, but is not a president seeking re-election.

<h2>The Data</h2>
I'll use data about the popular vote in each state since 1948 to determine whether incumbents performed consistently well or poorly in certain states over time, or whether their performance in different states was subject to the volatility in states (i.e. is incumbent performance consistent even in states considered "battleground" between parties?)

```{r, include = F}
library(dplyr)
library(maps)
library(ggplot2)
library(tidyverse)

state_votes <- read.csv("state_popvote_1948_2020.csv")
outcomes <- read.csv("popvote_1948-2020.csv") %>% filter(winner == TRUE)

# attach winner and incumbent status to each election in state_votes
state_votes <- state_votes %>%
               left_join(outcomes, by=join_by(year)) %>%
               mutate(state_winner=case_when(D_pv2p > R_pv2p ~ "DEM",
                                             R_pv2p > D_pv2p ~ "REP")) %>%
               group_by(state) %>%
               mutate(mean_D_pv2p=mean(D_pv2p, na.rm=T))

state_votes <- state_votes %>%
               mutate(state_voted_incumbent=case_when(party=="democrat" & incumbent_party==TRUE & state_winner=="DEM" ~ TRUE,
                                                   party=="republican" & incumbent_party==TRUE & state_winner=="REP" ~ TRUE,
                                                   party=="democrat" & incumbent_party==TRUE & state_winner=="REP" ~ FALSE,
                                                   party=="republican" & incumbent_party==TRUE & state_winner=="DEM" ~ FALSE,
                                                   party=="democrat" & incumbent_party==FALSE & state_winner=="DEM" ~ FALSE,
                                                   party=="republican" & incumbent_party==FALSE & state_winner=="REP" ~ FALSE,
                                                   party=="democrat" & incumbent_party==FALSE & state_winner=="REP" ~ TRUE,
                                                   party=="republican" & incumbent_party==FALSE & state_winner=="DEM" ~ TRUE))
```


<h2>Results</h2>

First, I decided to investigate whether particular states tend to vote for incumbent parties more often than others. The maps below show each state that voted for an incumbent party versus a challenger party's candidate in every election since 1948.

```{r, echo=F, warning = F}
# states where the incumbent won since 1948
states_map <- map_data("state")

state_votes %>%
  mutate(state = tolower(state)) %>%
  full_join(states_map, by=join_by(state == region)) %>%
  ggplot(aes(long, lat, group=group)) +
    facet_wrap(facets = year ~.) + 
    geom_polygon(aes(fill=state_voted_incumbent), color="white") +
    theme_void()
```

There is no visible trend of states that *consistently* vote for incumbent parties or *consistently* vote for challenger party candidates. So I decided to calculate the percentages of times since 1948 that each state has voted for an incumbent party's candidate. The results are in the table below:

```{r, echo=F, warning = F}
state_votes2 <- state_votes %>% group_by(state) %>% count(state_voted_incumbent) %>% filter(state_voted_incumbent==T) %>% select(state, n) %>% mutate(pct_incumbent=n/18) %>% arrange(desc(pct_incumbent))

data.frame(state_votes2)
```

Interestingly, the states that voted for an incumbent party *most* often candidate did so in two-thirds of the elections since 1948. Washington D.C. was *least* likely to pick an incumbent candidate, and it did so just over a third of the time since 1948. Overall, it makes sense that states do not *consistently* vote for an incumbent or a challenger over time because the incumbent's party tends to flip from Democrat to Republican, and vice versa, every few terms. Many states show consistency in the party they vote for, which means that they will necessarily swap from voting for to against an incumbent (or vice versa) every few elections. It would be most interesting to see if battleground states (states that notably *don't* show consistency in the party they vote for) tended to vote for incumbent parties more than challengers, but that is not evident from the analysis above.  
I was curious as to whether there is a relationship between partisanship and a state's tendency to vote for an incumbent party. The bar graph below shows how often states have voted for incumbent parties, colored by their average Democratic Vote Share since 1948.

```{r, fig.width=9, fig.height=9}
state_votes <- state_votes %>%
                left_join(state_votes2, by=join_by(state))

data <- state_votes %>% select(state, mean_D_pv2p, pct_incumbent) %>% distinct()

ggplot(data, aes(x=pct_incumbent, y=reorder(state, pct_incumbent), fill=mean_D_pv2p)) +
  geom_bar(stat="identity") +
  scale_fill_gradient2(low = "red", mid = "white", high = "navyblue", midpoint = 50, 
    breaks = 5, name = "Mean D Vote Share Since 1948") 
  
```

Based on the plot, it seems that some of the strongest Republican-leaning states tend to vote for incumbent candidates more frequently and that the more reliably Democratic states have voted for challengers more frequently. However, many of the most reliably-Democratic and reliably-Republican states have the same rates of voting for incumbents versus challengers (e.g. California and Idaho, and Rhode Island and Kansas, respectively).

<h2>How Should Campaigns Factor in Incumbency?</h2>
Professor Lynn Vavreck holds that, because the two main political parties in the United States have roughly equal support, they are less likely to adjust their ideology after an election than they are to try and adjust the rules of the electoral college or democratic system in their favor. They might be able to use figures about the states that tend to favor incumbents over challengers to inform the distribution of their campaign spending based on their party's status as an incumbent or challenger. However, the fact that the incumbent party changes every few years (perhaps a result of Incumbency Fatigue), makes it difficult to track differences in states' records voting for incumbent candidates versus challengers. States' tendencies to vote for a certain party, be it the incumbent party or not, are more salient.  


